Do investors in clean energy ETFs herd? The role of climate risks

### 2025-03-31 Journal/journal_submission/alt_energy_herding_anonymous.tex
comments

\title{Do investors in clean energy ETFs herd?}

\author{ 
Vassilios Babalos\footnote{Corresponding Author. Department of Accounting and Finance, University of Peloponnese, Kalamata,  Greece; Email: v.babalos@uop.gr.} \,\,
Xolani Sibande\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: xolaniss@gmail.com. Economic Research Department, South African Reserve Bank, Pretoria, South Africa.} \,\, 
Elie Bouri\footnote{Adnan Kassar School of Business, Lebanese American University, Lebanon; Email: elie.elbouri@lau.edu.lb} \,\,
Rangan Gupta\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: rangan.gupta@up.ac.za.} 
}
\date{\today}
\maketitle

@@

This study offers novel and valuable insights into herding behaviour in US clean energy ETFs between 2016 and 2023. 
The baseline herding tests by \cite{christie1995} and \cite{chang2000} revealed significant herding behaviour in this market. 
This evidence was supported by asymmetric and time-varying herding tests. That is, investors herd both in bear and bullish markets, and periodically. 
In addition, we found that climate risks (both physical and transition) reduced the probability of herding in US clean energy ETFs, 
indicating that an increase in climate-related risk encouraged efficient or climate-hedging behaviour by investors. 
Therefore, the results suggest that climate-related uncertainty did not drive herding behaviour in this market. The results suggest that investors are appropriately identifying opportunities that mitigate climate risk,
thereby reducing the probability of herding-driven system risk.

@@

\noindent\textbf{Keywords}: Herding Behaviour, Climate Change, Energy
\\
\textbf{JEL Codes}: G14, Q54, P18
\newpage

\setstretch{1.5}
\section{Introduction and literature
review}\label{introduction-and-literature-review}

Climate sustainability is a major concern for financial markets and
investors. The growing interest in climate sustainability is primarily
driven by the increasing materialisation of climate risks, and the
actions of governments, institutions and organizations towards a
sustainable future \citep{giglio2021a}. To ensure returns, investors
increasingly seek to hedge against climate risks by investing in green
financial products. Although the evidence is mixed, indications that
returns from green financial products are comparable to traditional
financial \citetext{\citealp[see amongst
others,][]{decclesia2024}; \citealp{nguyen2025}; \citealp{pastor2022}; \citealp[and][]{naqvi2022}}.

In this new climate sustainability paradigm, investors face many
pressures that not only bear on returns but also the stability of
financial markets. For example, resulting regulations aimed at reducing
emissions can surprisingly reduce the profitability of fossil-fuel-based
companies, or the possible mispricing of assets from ignoring climate
risks can lead to significant losses \citep{nguyen2025}. In addition to
these climate risks, a general change in investor attitudes can drive
the inclusion of green assets in their portfolio can lead to systemic
risk.\footnote{These pressures notwithstanding the possible contribution
  that financial markets can play in mitigating and reducing the
  negative effects of climate change \citep{giglio2021a}}

Specifically, climate change presents risks to investor portfolios
through two primary sources - physical and transition risks. Physical
risks or direct impact refer to extreme climate events such as floods
and droughts, which impact business operations and infrastructure; and
transition risks are the policy, technological, and other costs that
societies bear to achieve low carbon economies
\citetext{\citealp{nguyen2025}; \citealp[and][ amongst
others]{giglio2021a}}. Investors, therefore, recognise these risks and
seek to mitigate them, as they seek return-enhancing green financial
products.

Exchange-traded funds (ETFs) are a key feature of green financial
products. ETFs are a type of security that involves a collection of
securities that often tracks an underlying index. However, they can
invest in various industry sectors or strategies. In addition,
environment, social, and governance (ESG) ETFs serve as a market
discovery tool for investors to identify and invest in environmentally
friendly companies \citep{briere2023}. Among the ESG ETFs, the Clean
Energy (CE) ETFs have been the best-performing ones in 2022, followed by
the Cybersecurity and Artificial Intelligence (AI) ETFs
\citep{decclesia2024}. The clean energy transition represents one of the
largest multi-decade secular growth opportunities. After the inclusion
of Green energy financing in the list of United Nations Sustainability
Goals (SDGs) as SDG 7, the role, importance, and visibility of green
financial products have escalated enormously \citep{naqvi2022}. That is,
the growth of green assets under management is likely to continue.
Lastly, the limited availability of data on ESG complying investment
tools \citep{avramov2022, nguyen2025} justifies the use of green (CE)
ETFs as best candidates of green assets.

However, given how recent the inclusion of climate sustainability in
investment decisions is, it is not clear what the actual impact will be
in the long run. In this study specifically, we ask whether the rapid
adoption of CE ETFs could be driven by market fads, or is a fundamental
change in investor behaviour. Investors, for example, can believe that
peers have information about climate risks that they do not, investors
may herd to avoid losses compared to peers, or investors may be
encouraged to herd by the desire to align to climate-related social
values \citep{ciciretti2021, gavrilakis2023, loang2023}. Therefore,
market volatility and crisis, financial performance, and investor
sentiment can drive herding in ESG markets.

It is well established that herding literature is vast with
contradictory results depending mainly on the market, the employed
methodology and the period under consideration \citep{spyrou2013}.
Herding behavior can be either spurious in cases when investors make
similar decisions as a result of processing the same information set and
intentional herding when investors imitate the actions of others
\citep[see inter alia,][]{bikhchandani2000, galariotis2015}. Empirical
studies on herding usually fall into two categories: namely those that
employ holdings data aiming at measuring institutional investor herding
\citep{lakonishok1992}, and studies that use market returns data and
investigate herding towards the market consensus
\citep{chang2000, galariotis2015}. Our paper falls within the latter
category and tests for herding towards the market consensus for clean
energy US ETFs.

Herding behaviour in ESG markets is not without precedent. Amongst
others, \citet{loang2023} found that compliance with SDG goals can
introduce bias in investor sentiment, which leads to herding behaviour.
Using a Twitter (or X) uncertainty index, \citet{koutmos2024} found
evidence of herding in US-based ESG index fund investors.
\citet{przychodzen2016} found herding behaviour amongst fund managers
who incorporated ESG strategies in their portfolios. Lastly,
\citet{rubbaniy2021} highlight evidence of herding in the MSCI US ESG
Leader Index during extreme periods (bear and bull periods).

However, to the best of our knowledge, no studies exist that focus on
herding behaviour in CE ESGs. This study, therefore, aims to extend the
broader literature on herding behaviour in commodity and energy markets.
Several studies in this area were conducted. For example,
\citet{demirer2013a} conducted a commodity sectoral study and found
herding behaviour in grains but not in other sectors. Similarly,
\citet{gilbert2010} showed herding behaviour amongst speculators in
non-ferrous commodities. Others did not find evidence of herding in
similar markets. \citet{babalos2015} found significant anti-herding
behaviour in metal commodities futures after the global financial
crisis. \citet{pierdzioch2010} showed that forecasters in oil and metals
markets deviated from the crowd, indicating a rational response to
market information. \citet{steen2013} also found no herding behaviour in
international commodity markets. Overall, the literature in this area is
mixed, which indicates scope for further study.

In addition, our study extends the results of
\citet{dragomirescu-gaina2021} who examined herding behavior of
investors in the US energy sector and herding sensitivity to various
proxies of policy uncertainty and financial risk. They employed the
energy equities included in the S\&P 500 and concluded that herding
among investors in the US energy market sector is sensitive to green
volatility shocks

Therefore, this study investigates herding behaviour in alternative
energy ETFs in the US between 2016 and 2024. We then demonstrate how
climate-related uncertainty can drive herding behaviour in these
markets. Methodologically, we follow the standard herding tests by
\citet{christie1995} and \citet{chang2000}. The traditional approach was
supplemented by quantile regressions \citep{koenker1978} in order to
capture the time-varying aspects of herding. Lastly, we extend recent
approaches by \citet{bua2024} and others, which seek to establish a link
between climate uncertainty and herding behaviour.

@@

The sample consists of alternative energy equity ETFs (green ETFs) that
are traded in the US markets (see Table~\ref{tbl-data} in the
Appendix).\footnote{The data were sourced from
  \url{https://datastream.org/en-ca/}} The number of available
alternative energy ETFs in our sample varied from 10 in the beginning of
analysis to 30 at the most. The period of analysis runs from May
1\textsuperscript{st} of 2016 through 19\textsuperscript{th} June of
2024. The starting date was selected on the basis of the UN Climate
Change Conference (COP) Paris agreement. Daily logarithmic returns were
computed from the closing prices of ETFs for a total of 2122
observations.

@@

Following the relevant literature
\citetext{\citealp{christie1995}; \citealp[and][]{chang2000}}, we
compute dispersion of the \(i^{th}\) ETF from the market return. This is
known as the Cross Sectional Absolute Deviation (\(CSAD_t\)) measure.
Empirically the \(CSAD_t\) is defined in the following manner:

@@

Based on the above and in order to provide additional insight on the
herding phenomenon we examine whether herding presents an asymmetric
response on days when the market is up vis-à-vis days when the market is
down. To this end, we augment Equation~\ref{eq-herding} as follows:

@@

Rational asset pricing models \citep[for example,][]{black1972} predict
a linear relationship between return dispersion and market returns under
normal conditions, a relationship that is no longer valid in the
presence of herding. Herding behaviour leads to an increasing or
decreasing cross sectional dispersion with respect to market returns. In
other words, herding is captured by a non-linear term in the standard
pricing equation indicating a decreasing or an increasing returns'
dispersion.

Stated differently, as \citet{chang2000} argue, in the case of herding
the coefficient on the non-linear term (\(\gamma_2\)) will be negative
and statistically significant. Table~\ref{tbl-herding} presents the
results of herding for the full sample employing the non-linear
Equation~\ref{eq-herding}. The estimated coefficient on market return is
positive and highly significant as expected. The estimated coefficient
on the non-linear term is negative (-1.2773) and statistically
significant with a t-statistic of -9.71 suggesting that herd behaviour
is present and robust in the US alternative energy ETFs.

@@

There is ample evidence in the relevant literature that herding
behaviour in various asset markets \citep[see][]{pochea2017} exhibits
asymmetry and time-varying characteristics. To this end, we proceed to
estimate Equation~\ref{eq-herding} using the quantile regression (QR)
proposed by \citet{koenker1978}. Table~\ref{tbl-quantile_herding}
presents the results of estimating Equation~\ref{eq-herding} across
various quantiles of the returns dispersion. Our focus is on the herding
coefficient \(\gamma_2\), as a significant negative value of
\(\gamma_2\) is indicative of herding. Such a finding is observed at two
quantiles namely 25\% and 50\% with a value of -1.1056 and -1.165 which
are highly significant. It is worth mentioning that the sign of the
herding coefficient remains negative for almost all quantiles while the
significance changes from significant to insignificant while we move
from low and middle to upper quantiles (75\% and 90\%).

@@

It is widely accepted that asset returns are characterized by asymmetry,
that is, return dispersion tend to behave differently in rising and
falling markets \citep[see][]{geert2000, zhou2013, longin2001}. It
should be noted, that examining the relationship between returns
dispersion and market-wide returns across various quantiles of the
returns distribution allows us to make more robust inference regarding
the true behaviour of the phenomenon. Table~\ref{tbl-asymmetry_herding}
reports the estimation results of herding in the up and down markets
based on Equation~\ref{eq-asymmetry}. In general, we find that herding
is more likely to occur in down markets than in up markets, which is
indicative of the asymmetry of herding behaviour.

@@

\multicolumn{6}{>{\raggedright}m{\dimexpr 5.91in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{Note:\ This\ table\ presents\ the\ estimation\ results\ of\ herding\ of\ US\ Alternative\ energy\ equity\ ETFs\ according\ to\ Equation\ (3).\ *,**,***denotes\ significance\ at\ 10\%,5\%\ and\ 1\%\ respectively.}}}} \\

@@

Herding is present at all quantiles when markets are rising with an
estimated coefficient \(\gamma_3\) of -1.3669 and -1.2383 and highly
significant respectively. However, when markets are declining, investors
seem to neglect their own information set and imitate the actions of
others resulting in a highly significant coefficient of herding
(\(\gamma_4\)) across four out of five quantiles. Furthermore, we find
that in high quantiles (75\% and 90\%) and when markets are rising the
coefficient of interest (\(\gamma_3\)) turns positive but insignificant.

@@

There is ample evidence that herding might vary in response to market
conditions \citep[see][]{babalos2015, klein2013, stavroyiannis2019}. In
order to gain further insight on the time varying nature of herding we
conducted a rolling window analysis. The size of the rolling window is
related to the time-scales of the system (response times), and the aim
of the research \citep{babalos2015}. There is no golden rule for the
right size of the rolling window, there is a trade-off between having a
long enough window to estimate the metrics, and short enough to have a
sufficient number of windows in order to be able to derive a trend. In
light of the above discussion we set off to conduct a rolling window
analysis of 50 observations. Figure~\ref{fig-rolling_window} plots the
time evolution of the value of the estimated significance of the herding
coefficient (\(\gamma_2\)) using the rolling window analysis.

@@

We observe several periods of herding behaviour as reflected in the
troughs in Figure~\ref{fig-rolling_window}. The most prominent cases of
herding occur between March and May of 2020 followed by several
instances of herding in the period that extends from March through April
of 2017 (when \(\gamma_2\) averaged -45.5) and the period of
February-March of 2023 (when \(\gamma_2\) averaged -7.68).We observe
significant herding in the onset of the COVID-19 crisis 2020, which was
short lived as market adapted. The rest of the Covid period was in the
main characterised by anti-herding behaviour. \citet{ghorbel2023} and
\citet{dhall2020} found similar results. Lastly, similar to
\citet{mohamad2024} we observe no herding behaviour related to the
Russia-Ukraine war. \citet{mohamad2024} highlights the importance of
context in understanding the link between herding and market events. On
the other side, we derive significant moments of anti-herding behaviour
in the clean energy ETFs by observing the spikes in
Figure~\ref{fig-rolling_window}. Cross sectional dispersion appears to
increase with respect to market-wide returns which is a sign of
anti-herding behaviour on behalf of investors around December of 2016
and later during September of 2022.

\subsection{Climate-related uncertainty and herding
behaviour}\label{climate-related-uncertainty-and-herding-behaviour}

The behaviour of participants in energy markets is closely related to
the developments in the field of climate risks, carbon emissions and
environmentally friendly policies. In particular, following the
implementation of the Paris agreement in November 2016, climate policy
uncertainty has become in the epicenter of interest across carbon and
energy markets. There are a few studies that attempt to quantify the
effects of uncertainty related to climate on the economy and financial
markets \citep[see inter alia,][]{gabriel2024, bolton2021, krueger2020}.
To this end, \citet{bua2024} developed two climate risk related indexes
namely transition and physical risk using a text-based approach in order
to study the effect of these risks in financial markets. It is expected
that investors would prefer to hold assets that perform well in the face
of increasing climate change risks, even if this entails accepting lower
returns for such climate-hedging assets. Therefore, in the context of
our study and following previous studies that study the determinants of
herding behaviour \citep[see][]{bouri2019, demirer2018}, we attempt to
study the effect of climate-related uncertainty on the formation of
herding behavior in the clean energy market.

We use a probit model to relate herding to the two climate risk indexes
developed by \citet{bua2024} in the following manner:

@@

where \(D^{herd}\) takes a value of 1 during periods of statistically
significant herding (i.e., for days when the rolling t-statistic on
\(\gamma_2 < −1.96\) in Figure~\ref{fig-rolling_window}) and zero
otherwise. \(TRI\) is the transitional risk index and \(PRI\) is the
physical risk index.

The results from the Probit model are reported in
Table~\ref{tbl-probit_analysis}, where only the physical risk index
significantly decreases the probability of herding.\footnote{It should
  be noted that due to availability issues the probit analysis ends in
  December of 2023.} In other words, climate risks is good news for
clean energy stocks or firms resulting in anti-herding behaviour. This
implies that in the presence of higher physical risk with respect to the
climate, clean energy ETFs become a more attractive investment option
for investors that allocate their money to the various alternative
energy investment products. As a result, the cross-sectional dispersion
of clean energy ETFs tends to increase. Our results are in line with
those of relevant studies such as \citet{gabriel2024} who claimed that
in the event of climate policy shocks, clean energy assets could serve
the role of hedging instruments.

@@

Furthermore, we develop two additional models to study the effect of
climate risks on herding behaviour. We split the sample into two groups
based on the median value of the TRI and PRI. We then estimate the
following two models:

@@

We use these high PRI and high TRI in one probit regression and low TRI
and low PRI in another. Results are presented in
Table~\ref{tbl-probit_analysis2}. We observe higher uncertainty that
stems from physical or transition risk causes anti-herding which is in
line with the logic we discussed earlier.

@@

\renewcommand{\bibsection}{}
\bibliography{energy_herding.bib}

>>>>

\title{Do investors in clean energy ETFs herd? The role of climate risks}

%\author{ 
%Vassilios Babalos\footnote{Corresponding Author. Department of Accounting and Finance, University of Peloponnese, Kalamata,  Greece; Email: v.babalos@uop.gr.} \,\,
%Xolani Sibande\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: xolaniss@gmail.com. Economic Research Department, South African Reserve Bank, Pretoria, South Africa.} \,\, 
%Elie Bouri\footnote{Adnan Kassar School of Business, Lebanese American University, Lebanon; Email: elie.elbouri@lau.edu.lb} \,\,
%Rangan Gupta\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: rangan.gupta@up.ac.za.} 
%}
\date{\today}
\maketitle

@@

This study offers new insights into the herding behaviour in US clean energy Exchange-traded funds (ETFs) over the period from May 1, 2016 to June 19, 2024. 
The baseline model shows significant herding. An extended mode indicates that herding is present in both down and up markets, with a stronger effect in the down market, suggesting an asymmetry. 
Herding is also found to be time-varying. Further analysis shows that the transition climate risk, particularly its high level, reduces the probability of herding in clean energy ETFs, whereas physical climate risk does not exert any significant impact on the probability of herding. 
Thus, large levels of transition climate risk seem to encourage market efficiency in clean energy ETFs  and climate-hedging behavior by investors.

@@

\noindent\textbf{Keywords}: Herding Behaviour, Climate Change, Clean Energy
\\
\textbf{JEL Codes}: G14, Q54, P18
\newpage

\setstretch{1.5}
\section{Introduction}\label{introduction}

Climate sustainability is a major concern for financial markets and
investors. The growing interest in climate sustainability is primarily
driven by the increasing materialisation of climate risks, and the
actions of governments, institutions and organizations towards a
sustainable future \citep{giglio2021a}. To safeguard returns, investors
increasingly seek to hedge against climate risks by investing in green
financial products. Although the evidence from the existing literature
is mixed, returns from green financial products are somewhat comparable
to traditional financial products \citetext{\citealp[see amongst
others,][]{decclesia2024}; \citealp{nguyen2025}; \citealp{pastor2022}; \citealp[and][]{naqvi2022}}.
In this new climate sustainability paradigm, investors face many
pressures that not only bear on returns but also the stability of
financial markets. For example, resulting regulations aimed at reducing
emissions can surprisingly reduce the profitability of fossil-fuel-based
companies, or the possible mispricing of assets from ignoring climate
risks can lead to significant losses \citep{nguyen2025}. In addition to
climate risks, a general change in investor attitudes can drive the
inclusion of green assets in their portfolio can lead to systemic
risk.\footnote{These pressures notwithstanding the possible contribution
  that financial markets can play in mitigating and reducing the
  negative effects of climate change \citep{giglio2021a}} Specifically,
climate change presents risks to the global financial system, and
ultimately to investment portfolios, through two primary sources -
physical and transition risks. Physical risks or direct impact refer to
extreme climate events such as floods and droughts, which impact
business operations and infrastructure; and transition risks are the
policy, technological, and other costs that societies bear to achieve
low carbon economies \citetext{\citealp{nguyen2025}; \citealp[and][
amongst others]{giglio2021a}}. Investors, therefore, recognise these
risks and seek to mitigate them, as they seek return-enhancing green
financial products.

Exchange-traded funds (ETFs) are a key feature of green financial
products. They are a collection of investments that often tracks an
underlying performance of an asset or index. In the universe of
environment, social, and governance (ESG) investments, clean ETFs serve
as a tool for environmentally conscious investors to identify and invest
in environmentally friendly companies \citep{briere2023} engaged in the
transition towards a cleaner production and a low-carbon economy. Among
the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing
one in 2022 \citep{decclesia2024}. This is not surprising given that the
clean energy transition represents one of the largest multi-decade
secular growth opportunities. After the inclusion of Green energy
financing in the list of United Nations Sustainability Goals (SDGs) as
SDG 7, the role, importance, and visibility of green financial products
have escalated enormously \citep{naqvi2022}. That is, the growth of
green assets under management is likely to continue. Furthermore, the
limited availability of data on ESG complying investment tools
\citep{avramov2022, nguyen2025} justifies the use of energy ETFs as best
candidates of green assets. However, given that the inclusion of climate
sustainability in investment decisions is a recent phenomenon, it is not
clear what the actual impact will be in the long-run.

This study investigates herding behaviour in alternative (clean) energy
ETFs in the US between May 1 2016 and June 19 2024, showing evidence of
significant herding that is asymmetric and time-varying. We then study
whether climate-related uncertainty can affect the herding behaviour in
the US clean energy ETFs market. Methodologically, we follow the
standard herding tests by \citet{christie1995} and \citet{chang2000}.
Notably, we supplement the traditional approach with quantile
regressions \citep{koenker1978} in order to capture how herding differs
across various quantiles of the returns dispersion and in up and down
markets. Lastly, to establish a link between climate risk and herding
behaviour, we differentiating between transitional and physical risks
\citep{bua2024} and provide valuable insights on their potential impact
on the likelihood of herding in clean energy ETFs.

In this study specifically, we ask whether the rapid adoption of clean
energy ETFs could be driven by market fads, or is a fundamental change
in investor behaviour. Investors, for example, can believe that their
peers have more valuable information about climate risks, making them to
herd to avoid losses compared to peers; alternatively, investors may be
encouraged to herd by the desire to align to climate-related social
values \citep{ciciretti2021, gavrilakis2023, loang2023}. Investors in
clean energy stocks can experience large losses if their betting on
clean energy stocks go wrong, which represents a major concern in their
decision-making process. This incites those investors to disregard their
own information and follow the market consensus, leading to significant
herding behaviour in the clean energy market. \citet{devenow1996}
indicate that following the market consensus induces some kind of
security among less informed traders. This could be relevant to our
analysis because ETFs represent a popular investment instrument for
individual and retail investors who are not necessarily well informed
about the risk and prospects of investments in clean energy firms and
whether their contribution to the world transition to cleaner production
should be financially lucrative. On a related front, the
self-reinforcing nature of confidence in the tyranny of the majority, as
indicated by \citet{teraji2003} could also be pertinent.

While herding behaviour has been examined in ESG markets, it remains
understudied in clean energy assets, notably, clean energy
ETFs.\footnote{Amongst others, \citet{loang2023} shows that compliance
  with SDG goals can introduce bias in investor sentiment, which leads
  to herding behaviour. Using a Twitter (or X) uncertainty index,
  \citet{koutmos2024} finds evidence of herding in US-based ESG index
  fund investors. \citet{przychodzen2016} found indicate evidence of
  herding behaviour amongst fund managers who incorporated ESG
  strategies in their portfolios. Lastly, \citet{rubbaniy2021} highlight
  evidence of herding in the MSCI US ESG Leader Index during extreme
  (bear and bull periods) periods.} Other studies on herding have been
conducted in commodity and fossil energy markets. For example,
\citet{demirer2013} conducted a commodity sectoral study and found
herding behaviour in grains but not in other sectors. Similarly,
\citet{gilbert2010} shows herding behaviour amongst speculators in
non-ferrous commodities. Others did not find evidence of herding in
similar markets. \citet{babalos2015} find significant anti-herding
behaviour in metal commodities futures after the global financial
crisis. \citet{pierdzioch2010} show that forecasters in oil and metals
markets deviated from the crowd, indicating a rational response to
market information. \citet{steen2013} find no herding behaviour in
international commodity markets. Notably, our study extends the results
of \citet{dragomirescu-gaina2021} who have examined herding behavior of
investors in the US energy sector and herding sensitivity to various
proxies of policy uncertainty and financial risk. They study the energy
equities included in the S\&P 500 and concluded that herding among
investors in the US energy market sector is sensitive to green
volatility shocks

Our analysis shows that herding is significant and is present in both
down and up markets, with a stronger effect in the down market,
suggesting an asymmetry. Herding is also found to be time-varying.
Notably, an additional analysis reveals that the transition climate
risk, particularly its high level, reduces the probability of herding in
clean energy ETFs, whereas physical climate risk does not exert any
significant impact on the probability of herding

@@

The sample consists of clean energy equity ETFs (green ETFs) that are
traded in the US markets (see Table~\ref{tbl-data} in the
Appendix).\footnote{The data were sourced from
  \url{https://datastream.org/en-ca/}} The number of available clean
energy ETFs in our sample varied from 10 in the beginning of analysis to
30 at the most. The period of analysis runs from May 1 2016 to 19 June
2024. Daily closing prices on the clean energy ETFs under study were
collected from the Refinitiv database. The starting date was selected on
the basis of the UN Climate Change Conference (COP) Paris agreement.
Daily logarithmic returns were computed from the closing prices of each
ETF, yielding, a total of 2122 observations per ETF.

@@

It is well established that herding literature is vast with
contradictory results depending mainly on the market, the employed
methodology and the period under consideration \citep{spyrou2013}.
Herding behavior can be either spurious in cases when investors make
similar decisions as a result of processing the same information set and
intentional herding when investors imitate the actions of others
\citep[see inter alia][]{bikhchandani2000, galariotis2015}. Empirical
studies on herding usually fall into two categories: namely those that
employ holdings data aiming at measuring institutional investor herding
\citep[e.g.][]{lakonishok1992}, and studies that use market returns data
and investigate herding towards the market consensus
\citep[e.g.][]{chang2000, galariotis2015}. Our paper falls within the
latter category and tests for herding towards the market consensus for
clean energy US ETFs.

Following the relevant literature
\citetext{\citealp{christie1995}; \citealp[and][]{chang2000}}, we
compute dispersion of the \(i^{th}\) ETF from the market return, which
is known as the Cross Sectional Absolute Deviation (\(CSAD_t\)) measure.
Empirically the \(CSAD_t\) is defined in the following manner:

@@

To provide additional insight on the herding phenomenon we examine
whether herding presents an asymmetric response on days when the market
is up vis-à-vis days when the market is down. To this end, we augment
Equation~\ref{eq-herding} as follows:

@@

Rational asset pricing models \citep[for example,][]{black1972} predict
a linear relationship between return dispersion and market returns under
normal conditions, a relationship that is no longer valid in the
presence of herding. Herding behaviour leads to an increasing or
decreasing cross sectional dispersion with respect to market returns. In
other words, herding is captured by a non-linear term in the standard
pricing equation indicating a decreasing or an increasing returns'
dispersion. Stated differently, as \citet{chang2000} argue, in the case
of herding the coefficient on the non-linear term (\(\gamma_2\)) will be
negative and statistically significant.

Table~\ref{tbl-herding} presents the results of herding for the full
sample, based on Equation~\ref{eq-herding}. The estimated coefficient on
market return is positive and highly significant as expected.
Importantly, the estimated coefficient on the non-linear term is
negative (-1.2773) and statistically significant with a t-statistic of
-9.71 suggesting that herd behaviour is present and robust in the US
alternative energy ETFs. Accordingly, investors in US clean energy ETFs
tend to disregard their private information and follow market consensus.

@@

Evidence from the existing literature shows that herding behaviour in
various asset markets \citep[see][]{pochea2017} exhibits asymmetry. To
this end, we proceed first with the estimation of
Equation~\ref{eq-herding} using the quantile regression (QR) proposed by
\citet{koenker1978}. Table~\ref{tbl-quantile_herding} presents the
estimated results across various quantiles of the returns dispersion.
They show that herding is statistically significant at lower
(\(\tau= 25\%\)), middle (\(\tau =50\%\)), and upper (\(\tau=75\%\))
with a value of -1.1056, -1.165, and -1.1473, respectively. No
significant herding is found at extreme lower quantile (\(\tau=10\%\))
and extreme higher quantile (\(\tau=90\%\)).

@@

Then, we study herding in up and down markets. It is widely accepted
that asset returns are characterized by asymmetry, that is, return
dispersion tend to behave differently in rising and falling markets
\citep[see][]{geert2000, zhou2013, longin2001}, with evidence suggesting
that herding is more pronounced during periods of market stress. In this
regard, examining the relationship between returns dispersion and
market-wide returns across various quantiles of the returns distribution
of clean energy ETFs allows us to make more robust inference regarding
the true behaviour of the herding phenomenon.
Table~\ref{tbl-asymmetry_herding} reports the estimation results of
herding in the up and down markets based on Equation~\ref{eq-asymmetry}.
In general, we find that herding is significant in both down and up
markets

@@

\multicolumn{6}{>{\raggedright}m{\dimexpr 5.91in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{Note:\ This\ table\ presents\ the\ estimation\ results\ of\ herding\ of\ US\ Alternative\ energy\ equity\ ETFs\ according\ to\ Equation\ (3).\ }}}\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{$\gamma_3$}}}\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\ captures\ potential\ herding\ in\ the\ up\ market,\ whereas\ }}}\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{$\gamma_4$}}}\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\ captures\ potential\ herding\ in\ the\ down\ market.\ *,**,***denotes\ significance\ at\ 10\%,5\%\ and\ 1\%\ respectively.}}}} \\

@@

When the market is rising, herding is present at all quantiles with a
statistically significant coefficient \(\gamma_3\) ranging from -0.2522
(\(\tau=10\%\)) to -0.2942 (\(\tau=90\%\)) . Similarly, when markets are
declining, investors neglect their own information set and imitate the
actions of others resulting in a highly significant coefficient of
herding (\(\gamma_4\)) across four out of five quantiles, ranging from
-0.9721 to -1,3144. Notably, the coefficient of herding in down market
is larger than in up market, reflecting an asymmetric herding behaviour.

@@

There is ample evidence that herding varies with time
\citep[see][]{babalos2015, klein2013, stavroyiannis2019} and intensifies
during crisis periods. In order to gain insights on the time-varying
nature of herding in clean energy ETFs, we conduct a rolling window
analysis. The size of the rolling window is related to the time-scales
of the system (response times), and the aim of the research
\citep{babalos2015}. There is no golden rule for the right size of the
rolling window, there is a trade-off between having a long enough window
to estimate the metrics, and short enough to have a sufficient number of
windows in order to be able to derive a trend. Accordingly, we conduct a
rolling window analysis of 50 observations, and plot in
Figure~\ref{fig-rolling_window} the time evolution of the value of the
estimated significance of the herding coefficient (\(\gamma_2\)).

@@

We observe several periods of herding behaviour as reflected in the
troughs in Figure~\ref{fig-rolling_window}. The most prominent cases of
herding occur between March and May of 2020 followed by several
instances of herding in the period that extends from March through April
of 2017 and the period of February-March of 2023. On the other side, we
derive significant moments of anti-herding behaviour in the clean energy
ETFs by observing the peaks in Figure~\ref{fig-rolling_window}. Cross
sectional dispersion appears to increase with respect to market-wide
returns, which is a sign of anti-herding behaviour on behalf of
investors around December of 2016 and later during September of 2022

\subsection{Climate risks and herding
behaviour}\label{climate-risks-and-herding-behaviour}

The behaviour of participants in energy markets is closely related to
the developments in the field of climate risks, carbon emissions, and
environmentally friendly policies. Rising climate risk is found to
increase green energy prices \citep{dutta2023}, and evidence from the
existing literature shows that climate policy uncertainty affects the
performance of clean energy stocks relative to dirty ones, making the
former outperform the later when the levels of climate policy
uncertainty are high \citep{bouri2022}. In particular, following the
implementation of the Paris agreement in November 2016, climate policy
uncertainty has become in the epicenter of interest across carbon and
energy markets. There are a few studies that attempt to quantify the
effects of uncertainty related to climate on the economy or financial
markets \citep[see inter alia,][]{gabriel2024, bolton2021, krueger2020}.
Interestingly, \citet{bua2024} developed two climate risk related
indexes namely transition and physical risk using a text-based approach
in order to study the effect of these risks in financial markets. In
this regard, \citet{bouri2023} study the impact of both physical and
climate risks on the returns and volatility of brown and green energy
stocks, carbon emission allowances, and green bonds, showing evidence
that transitional climate risk exerts a more significant impact than
physical climate risk.

It is expected that environmentally conscious investors would prefer to
hold clean energy assets that perform well in the face of increasing
climate change risks \citep[see][]{bouri2022}, even if this entails
accepting lower returns for such climate-hedging assets. Therefore, in
the context of our study and following previous studies on the
determinants of herding behaviour \citep[see][]{bouri2019, demirer2018},
we examine the effect of physical and transition climate risks on the
formation of herding behavior in the clean energy ETF market.

To this end, we use a probit model to relate herding to the two climate
risk indexes developed by \citet{bua2024} in the following manner:

@@

where \(D^{herd}\) takes a value of 1 during periods of statistically
significant herding (i.e., for days when the rolling t-statistic on
\(\gamma_2 < −1.96\) in Figure~\ref{fig-rolling_window}) and zero
otherwise. \(TRI\) is the transitional risk index and \(PRI\) is the
physical risk index. For details on the construction of \(TRI\) and
\(PRI\), the reader can refer to the paper of \citet{bua2024}.

The results from the Probit model are reported in
Table~\ref{tbl-probit_analysis}. Showing that only the transitional
climate risk index significantly decreases the probability of herding in
clean energy ETFs.\footnote{It should be noted that due to the
  availability of climate risk data from \citet{bua2024}, the probit
  analysis covers the period from May 1, 2016 to December 30, 2023.}
Transitional climate risk represent good news for investors in clean
energy stocks, possibly reducing their self-reinforcing nature of
confidence in the tyranny of the majority \citep{teraji2003}) and their
need for shared intention and action, resulting in a decrease in the
herding behaviour. In the presence of higher transitional risk with
respect to the climate, clean energy ETFs become a more attractive
investment alternative for environmentally conscious investors who
allocate their money to alternative energy investment products
\citep[see][]{bouri2022}, reinforcing the confidence of investors in
their own information. As a result, the cross-sectional dispersion of
clean energy ETFs tends to increase. Our results are somewhat in line
with \citet{bouri2023} who show that the transitional climate risk is
more important than physical risk for the return and volatility of clean
energy stocks. They also concord with other relevant studies which
indicate that in the event of climate policy shocks, clean energy assets
could serve the role of hedging instruments \citep{gabriel2024} and tend
to outperform brown energy assets \citep{bouri2022}.

@@

Furthermore, we develop two additional models to study the effect of
high and low levels of climate risks on herding behaviour in clean
energy ETFs. Accordingly, we split the sample into two groups based on
the median value of the \(TRI\) and \(PRI\) and estimate the following
two models:

@@

Using these high \(PRI\) and high \(TRI\) in one probit regression and
low \(TRI\) and low \(PRI\) in another one, we present the results in
Table~\ref{tbl-probit_analysis2}. We observe that high levels of
transition risk decrease the likelihood of herding (i.e.~drives
anti-herding) at the 10\% level of significance, which is in line with
the logic we discussed earlier.

@@

\renewcommand{\bibsection}{}
\bibliography{references.bib}
